%
%	GENEsYs-rep.tex		Ba, 22 feb 91		LaTeX 2.09
%				Ba, 15 jun 92
%
%	The report file and its use as well as the other files.
%{

\section{The Report}\label{genesis-rep}

A report describing the performance of the \GA\ is automatically
produced by the ``report'' program after a run of the \GA.
The report summarizes the mean,  variance and  range of several 
measurements, including online performance, offline performance, 
the average performance  of  the current population, and the 
current best value\footnote{%
	I.e.~the best value of the actual generation. 
	In \Ogen\ the all--time best value was reported, 
	which seems not appropriate, since we are interested in informations 
	about the actual population.}.  
Online performance is the mean of all  evaluations.   
Offline  performance  is  the  mean  of  the current best evaluations  
(see~\cite{Jon80a}).  
\Vrb{Conv} indicates the number of positions in which at least  $95\%$  of
the  population  has  the  same value.  
\Vrb{Lost} indicates the number of positions in which $100\%$ of the 
population has  the same  value.  
\Vrb{Bias} indicates the average percentage of the most prominent value in 
each position.  
That is, a  bias  of $0.75$  means that, on average, each position has 
converged to $75\%\ 0$'s or $75\%\ 1$'s.  
The minimum bias is $0.50$.
If adaptive mutation rates have been selected, the file additionally contains
the average, maximum, and minimum mutation rate contained in the population.
A typical \File{rpt}-file looks like the following:

\begin{scriptsize}
\begin{center}
\begin{tabular}{rrcccrrrrr}
\ifUS
	\multicolumn{2}{c}{{\sc Mean}}&&&&&&&&\\
Gens & Trials &Lost &Conv  &Bias   &Online  & Offline   &  Best  &  Average &  Worst \\
   0   &  50  &  0  &  0 &0.568 &2.74e+01 &7.28e+00 &2.00e+00 &2.74e+01 &5.64e+01	\\
   3   & 200  &  0  &  0 &0.607 &2.06e+01 &2.77e+00 &6.93e-01 &1.41e+01 &3.21e+01	\\
   7   & 400  &  0  &  1 &0.658 &1.51e+01 &1.73e+00 &6.93e-01 &6.79e+00 &1.92e+01	\\
  11   & 600  &  2  &  3 &0.681 &1.15e+01 &1.35e+00 &1.89e-01 &3.52e+00 &1.75e+01	\\
  15   & 800  &  2  &  5 &0.701 &9.32e+00 &1.05e+00 &1.60e-01 &2.05e+00 &8.36e+00	\\
  19   &1000  &  3  &  6 &0.723 &7.77e+00 &8.76e-01 &1.57e-01 &1.18e+00 &3.22e+00	\\
\else
 {\sc Mean}&&&&&&&&&\\
Gens & Trials &Lost &Conv  &Bias   &Online  & Offline   &  Best  &  Average &  Worst \\
   0   &  50  &  0  &  0 &0.568 &2.746e+01 &7.289e+00 &2.002e+00 &2.746e+01 &5.644e+01	\\
   3   & 200  &  0  &  0 &0.607 &2.065e+01 &2.771e+00 &6.934e-01 &1.418e+01 &3.217e+01	\\
   7   & 400  &  0  &  1 &0.658 &1.510e+01 &1.732e+00 &6.934e-01 &6.791e+00 &1.925e+01	\\
  11   & 600  &  2  &  3 &0.681 &1.156e+01 &1.350e+00 &1.894e-01 &3.525e+00 &1.758e+01	\\
  15   & 800  &  2  &  5 &0.701 &9.326e+00 &1.056e+00 &1.602e-01 &2.053e+00 &8.365e+00	\\
  19   &1000  &  3  &  6 &0.723 &7.775e+00 &8.765e-01 &1.572e-01 &1.188e+00 &3.222e+00	\\
\fi
\end{tabular}
\end{center}
\end{scriptsize}

In this file, only average values over the given number of experiments are
collected.
If more than one experiment was performed, the variances of these measures 
are also calculated and collected in a file named \File{var}.
The structure of the file is completely identical to the structure of the
\File{rep}-file.

%}
